Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-27 manuscript.tex
latest

Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients. This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory. Following in the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, but grounding ourselves empirically in ``primary strategy'' trait classification schemes like Grime's triangle, we develop a new model of density-dependent selection which revolves around territorial contests. To do so, we generalize the classic lottery model of territorial acquisition, which has primarily been used for studying species co-existence questions, to accommodate arbitrary densities. We use this density-dependent lottery model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal scheme. We revisit previous concepts of density-dependent selection, including $r$ and $K$ selection, and argue that our model distinguishes between different aspects of fitness in a more natural and intuitive manner.

@@

``...the concept of fitness is probably too complex to allow of a useful mathematical development. Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J. Ewens, Mathematical Population Genetics I, 2004

@@

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a genotype with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two genotypes). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{1}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}c_jl_j}.\label{eq:Da}
\end{equation}

@@

Fig.~\ref{fig:simcomp} shows that Eq. \eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities. Two genotypes are present, one of which is at low frequency. The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b). On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d). Fig. 3 also shows the breakdown of the classic lottery model at low propagule densities.

@@

%From Eq. \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates. For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested. The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).

@@

The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row; details in Appendix). These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.

According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate, with a proportionality factor typically of order one (\citealt{uecker_2011}) [find Haldane reference]. The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit, where availability is subject to constraints imposed by the environment.

@@

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b\gg d$ & $b\gg d$ & $b\approx d$ & $b>d$ \\
  Density $N/T$  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment. Shown are the mapping of each environment to our parameters, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

In the previous section we only considered the how $b$, $c$ and $d$ should respond in Grime's environmental extremes. Here we further explore the low frequency behavior of Eq. \eqref{eq:master} to determine which types can coexist in a constant environment, and then consider the full time-dependent behaviour of Eq. \eqref{eq:master} in a cyclical environment.

@@

This classical argument overlooks two effects that are present in our extended lottery model. The first is the storage effect, which requires part of the population to be protected from selection each iteration (as occurs when generations are non-overlapping) \citep{chesson_1981}, and which is also present in the classic lottery model. The second is the  bounded population size effect of \cite{yi_2013}, which requires fluctuations in population density and density-dependent regulation of population growth.

@@

Figure \ref{fig:fluctuatingselection} shows the behavior of Eq. \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth). Both the storage effect (adults are sheltered from selection during the summer growth phase) and the bounded density effect (expansion to high density occurs every cycle) are operating. Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$).

Neither type has an advantage over a full environmental cycle, and they stably coexist. This is due to some combination of the storage and bounded population size effects (stable coexistence between $b$ and $d$ specialists was not possible in a constant environment). The classic lottery model gives dramatically different co-existence predictions in this scenario (keeping all other parameters the same, the birth rate of the winter survivor must be about $5$ times smaller). This occurs because density returns immediately to capacity $N=T$ in the first summer iteration, after which type frequencies remain constant until the winter. The $d$ specialist thus effectively has infinitely many propagules to secure its winter frequency gains, an enormous advantage compared to the finite propagule density dynamics in Fig. \ref{fig:fluctuatingselection}. Similar difficulties arise in previous models of how the storage effect promotes genetic variation \citep{ellner_1994}, which assume that the total number of offspring per iteration is constant. Beyond this observation, disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here. Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density dependent effects.

\begin{figure}
\centering
\includegraphics[scale=0.7]{fluctuatingselection.pdf}
\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment. (a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer). The $b$ specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$ specialist ($b=0.2$, $d=0.1$) (blue). (b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist does so at a lower rate. Total height (blue+black) is population density $N/T$. (c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist. For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed disperal.} 
\end{figure}

@@

In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference. While we have not directly attempted to perform inference with our model, and Eq. \eqref{eq:master} may not be appropriate for particular inference problems, it seems that something similar (and hopefully more analytically tractable) to our extension of the lottery model is unavoidable because, fundamentally, selective births and deaths affect both  abundances and frequencies, not one or the other in isolation. Moreover, some aspects of allele frequency change are intrinsically density dependent. In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model''). This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist. This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}. It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}). In cyclical environments, polymorphisms can be stablized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}. We leave thedetails of how our model might be applied in inference, including the crucial issue of its genetic drift predictions, for future work.

It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme. While adaptive evolution in the direction predicted by our model does produce traits broadly consistent with Grime's scheme, trait data on selection for higher $b$ at high density was ambiguous. This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection. Yet it is not surprising that $b$ can matter at high densities. In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable. This is a realistic feature of the model. Even if one genotype is guaranteed to win a territory in a ``fair'' contest (e.g. it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance. For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental advantage. First arrivals are more likely to occur for genotypes with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models. The analogous intuition in the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though population size is not changing. The logistic model for which $r$ and $K$ are named, does not capture this intuition.

Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}. Contrary to an $r$/$K$ dichotomy, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.

There is support for a negative relationship between competitive success at high density and maximal growth rate \citep{luckinbill_1979}, consistent with an $r$/$K$ dichotomy. This could be driven by a tradeoff between individual size and reproductive rate. To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}. However, competitive success as measured by $\alpha$ (i.e. the per-capita effect of one genotype on another genotype's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each genotype to produce contestants i.e. $b$ in our model. Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (the total number of territories occupied in an iteration is $\Delta_+ N=U(1-e^{-L})$, which does not depend directly on the $c_i$). The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum relative fitness trait within a non-zero-sum fitness model. This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to clonal interference \citep{gerrish_1998,desai_2007} between $c$-strategists on the one hand, and $b$- and $d$- strategists on the other.

@@

While our model can be applied to species rather than genotypes (e.g. ecological invasions), our focus is genotype evolution. Our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ in the population. Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases. However, this restriction could become important when looking at species interactions rather than genotype evolution.

@@

%\subsection{Logistic growth and classic nesting sites model}

%In the nesting sites model, it is assumed that propagules landing on occupied territory do not survive, while those landing on unoccupied territory immediately occupy the territory. Thus, the probability that a propagule survives is simply $(1-N/K)$, where $N=\sum_i n_i$ is the total population size. Consequently, $n_i$ increases by $1$ over the interval $[t,t+\Delta t]$ with probability $b_i n_i (1-N/K)\Delta t$. Provided that the abundances $n_i$ are large enough that demographic stochasticity is negligible (i.e. they have established), we can treat $n_i(t)$ as a continuous variable with growth given by the expectation of $b_i n_i (1-N/K)\Delta t$, and so, taking $\Delta t \rightarrow 0$ we obtain
%\begin{equation}
%\frac{dn_i}{dt}=b_i\left(1-\frac{N}{K}\right)n_i. \label{eq:logistic}
%\end{equation}

%\section{Territorial availability}
%
%In Sec. \ref{sec:c} we assumed that adults have the same territorial requirement, regardless of genotype. This can be generalized to allow genotypes to differ in their territorial requirements. Each adult from genotype $i$ now requires $t_i$ units of territory, $T$ is the total territory available to the population, and $T-\sum_i t_i$ units of territory are unoccupied. 
%
%Competition between propagules with different territorial requirements is potentially much more complicated than the model in Sec. \ref{sec:c}, because competition is no longer neatly partitioned into a set of identical territories.  The larger a propagule's territorial requirement, the more neighboring propagules it will interact with during its development to adulthood. 
%
%
%A simple way to avoid these complications is to partition the unoccupied territory into uniform units of territory with size given by the largest territorial requirement $t^*={\rm max}_i t_i$. The uniform territory approach used in Sec.  \ref{sec:c}
%
%The biological intuition behind this 
%
%Thus $U=(T-\sum_i t_i n_i)/t^*$.

>>>>

[Needs to be revised but won't do that now in case it has to become Nature introductory paragraph]
Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients. This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory. Following in the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, but grounding ourselves empirically in ``primary strategy'' trait classification schemes like Grime's triangle, we develop a new model of density-dependent selection which revolves around territorial contests. To do so, we generalize the classic lottery model of territorial acquisition, which has primarily been used for studying species co-existence questions, to accommodate arbitrary densities. We use this density-dependent lottery model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal scheme. We revisit previous concepts of density-dependent selection, including $r$ and $K$ selection, and argue that our model distinguishes between different aspects of fitness in a more natural and intuitive manner.

@@

``...the concept of fitness is probably too complex to allow of a useful mathematical development. Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J. Ewens, Mathematical Population Genetics I, 2004 [Should be removed if we need space]

@@

Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$. For a genotype with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$. To account for this, we separate Eq. \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts. Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two genotypes). We obtain (details in Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{1}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}c_jl_j}.\label{eq:Da}
\end{equation}
To supplement our analytical mean field derivation, we verified that Eq. \eqref{eq:master} closely approximates the exact behavior of our density-dependent lottery model using numerical simulations (Appendix B). Thus, Eq. \eqref{eq:master} describes how type abundances change over time in a lottery model where population density can itself vary with time.

@@

@@

@@

The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row; details in Appendix). These can be used to infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.

According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}, with a proportionality factor typically of order one. The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). The direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit \citep{yampolsky_2001}, where availability is subject to constraints imposed by the environment.

@@

\begin{figure}
\centering
\begin{tabular}{l*{4}{c}}
  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
  Other parameters & $b\gg d$ & $b\gg d$ & $b\approx d$ & $b>d$ \\
  Density $N/T$  & High & Low & Low & High \\
  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment. Shown are the mapping of each environment to our parameters, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$. *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq. \eqref{eq:delttot}.}
\end{figure}

@@

In the previous section we only considered how $b$, $c$ and $d$ should respond in Grime's environmental extremes. Here we further explore the low frequency behavior of Eq. \eqref{eq:master} to determine which types can coexist in a constant environment, and then consider the full time-dependent behaviour of Eq. \eqref{eq:master} in a cyclical environment.

@@

This classical argument overlooks two general effects that promote coexistence in fluctuating environments. The first is the storage effect, which occurs when part of the population is protected from selection (due to overlapping generations in the lottery model; \citealt{chesson_1981}). The second is the  bounded population size effect of \cite{yi_2013}, which occurs when each environmental cycle involves growth from low to high density, with the time spent growing each cycle dependent on the fitness of the types present.

@@

Fig. \ref{fig:fluctuatingselection}a-c shows the behavior of Eq. \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth). Both the storage effect (adults are sheltered from selection during the summer growth phase) and the bounded density effect (expansion to high density occurs every cycle) are operating. Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$). Neither type has an advantage over a full environmental cycle, and they stably coexist. This is due to a combination of the storage and bounded density effects (recall that stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).

The classic lottery model (Eq. 1) gives dramatically different co-existence predictions because expansion to carrying capacity occurs immediately at the start of the summer (Fig. \ref{fig:fluctuatingselection}d-f). As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than predicted when we properly account for the growth in the abundance of each type using Eq. \eqref{eq:master} (keeping the other parameters the same). Previous models of the promotion of genetic variation via the storage effect  \citep{ellner_1994} similarly assume that the total number of offspring per iteration is constant, and would produce a similar error.

%Disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here. Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density-dependent effects.

\begin{figure}
\centering
\includegraphics[scale=0.7]{fluctuatingselection.pdf}
\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment. (a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer). The $b$ specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$ specialist ($b=0.217$, $d=0.1$) (blue). (b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist does so at a lower rate. Total height (blue+black) is population density $N/T$. (c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist. (d-f) Same as (a-c) for the classic lottery model, where coexistence now requires $b\approx 0.0421$ for the $d$ specialist. For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed dispersal.} 
\end{figure}

@@

In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference. While we have not directly attempted to perform inference with our model, and Eq. \eqref{eq:master} may not be appropriate for particular inference problems, it seems that something similar (and hopefully more analytically tractable) to our extension of the lottery model is unavoidable because, fundamentally, selective births and deaths affect both  abundances and frequencies, not one or the other in isolation. Moreover, some aspects of allele frequency change are intrinsically density-dependent. In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model''). This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist. This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}. It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}). In cyclical environments, polymorphisms can be stabilized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}. We leave the details of how our model might be applied in inference, including the crucial issue of its genetic drift predictions, for future work.

It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme, where $r=b-d$ is the maximal, low-density growth rate \citep{pianka_1972}. Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}. Contrary to an $r$/$K$ dichotomy, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.

%While adaptive evolution in the direction predicted by our model does produce traits broadly consistent with Grime's scheme, trait data on selection for higher $b$ at high density was ambiguous. This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection. Yet it is not surprising that $b$ can matter at high densities. In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable. This is a realistic feature of the model. Even if one genotype is guaranteed to win a territory in a ``fair'' contest (e.g. it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance. For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental advantage. First arrivals are more likely to occur for genotypes with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models. The analogous intuition in the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though population size is not changing. The logistic model for which $r$ and $K$ are named, does not capture this intuition.

There is support for a negative relationship between competitive success at high density and maximal growth rate \citep{luckinbill_1979}, consistent with an $r$/$K$ dichotomy. This could be driven by a tradeoff between individual size and reproductive rate. To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}. However, competitive success as measured by $\alpha$ (i.e. the per-capita effect of one genotype on another genotype's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each genotype to produce contestants i.e. $b$ in our model. Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (the total number of territories occupied per iteration is $\Delta_+ N=U(1-e^{-L})$, which does not depend directly on the $c_i$). The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum relative fitness trait within a non-zero-sum fitness model. This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to clonal interference \citep{gerrish_1998,desai_2007} between $c$-strategists on the one hand, and $b$- and $d$- strategists on the other.

@@

While our model can be applied to species rather than genotypes (e.g. ecological invasions), our focus is genotype evolution i.e. the change in allele frequencies over time. Our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ in the population. Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases. However, this restriction could become important when looking at species interactions rather than genotype evolution.

@@

\subsection*{Comparison with simulations}

Fig.~\ref{fig:simcomp} shows that Eq. \eqref{eq:master} and its components closely approximate our density-dependent lottery model over a wide range of propagule densities (the latter is evaluated by direct simulations of uniform random dispersal and lottery competition). Two genotypes are present, one of which is at low frequency. The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b). On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d). Fig. 3 also shows the breakdown of the classic lottery model at low propagule densities.

\section*{Appendix C: Direction of evolution under different environmental extremes}

Summing Eq. \eqref{eq:master} over genotypes, $N$ obeys 
\begin{equation}
\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
\end{equation}
where $U=\sum_i b_i n_i/L$. In equilibrium ($\Delta N=0$), Eq. \eqref{eq:deltN} implies $L/(1-e^{-L})=\sum_i b_i n_i/\sum_i d_i n_i$, and so $L$ only depends on the ratio of population mean birth rate to population mean death rate.

and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested. The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).
